---
title: "Linear Regression Statistical Analysis Project "
author: "Aviv Gelfand and Maor Moshe  -  03/12/2022"
output:
  html_document: 
    fig_height: 4
    df_print: kable
    fig_width: 6
  word_document: default
editor_options:
  chunk_output_type: console
---
### 1) About the Data-Set
We used WHO table of Average BMI and life expectancy for 177 countries in the world.
[Source](https://www.kaggle.com/datasets/kumarajarshi/life-expectancy-who?resource=download).

**Predicted / dependent variable Y**: Life expectancy - refers to average number of years a person can expect to live in a country.

**Undependable variable X**: BMI - Body Mass Index units. A calculation using a person's height and weight. The formula is $BMI = \frac{kg}{h_m^2}$ where kg is a person's weight in kilograms and mh is their height in meters. A BMI of 25.0 or more is overweight, while the healthy range is 18.5 to 24.9.

**We hypothesize** that BMI is an explanatory variable for a person's life expectancy. 

##### 1 i) Handling an anomaly in the data
We noticed there are observations between 0 and 10 (physically impossible BMI values). WE Removed the noise in the data. Observations where BMI is smaller then 10 are omitted.

```{r echo=FALSE, message=FALSE, warning=FALSE }
# getwd()
setwd("C:/Users/avivg/OneDrive/Documents/R/Statistical Analysis Course/Lab")
df <- read.csv("WHO_BMI_2015.csv")
Life.expectancy <- df$Life.expectancy
BMI <- df$BMI
# head(df)
# na.omit(df)
plot(main = "Clean Scatter Plot",BMI,Life.expectancy, col="blue")
# abline(v=10, col="red") 
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
#BMI1 <-  df$BMI >10
#df2 <- df[df$BMI > 10,]
#df <- df2
#plot(df2$BMI, df2$Life.Expectancy.Years )
```


### 2) Discriptive Statistics
#### Histograms:
```{r}
hist(BMI , main = "BMI", breaks=50, freq=FALSE, col="light blue")
lines(density.default(BMI))
```

A distribution with two centric values can be seen(23,27).no values under 20 or above 32. (that indicates precise data because beyond this range its means extreme underweight or extreme obesity) 

```{r}
hist(Life.expectancy, main = "Life Expectancy", breaks=50, freq = FALSE, col="light blue")
lines(density.default(Life.expectancy))
```

It can be seen that the life expectancy dis' has a long left tail and a center around `r round(mean(Life.expectancy))`, and the latest age is `r round(max(Life.expectancy))`

You can see observations that represent an average life expectancy below the age of 55. We assume that these observations will correspond with people with a low BMI score - third world countries. We will confirm the test later in the work. 

##### Mean, Median, SD, Min, Max,  and Quantiles:

```{r}
# tmp <- df[c('BMI','Life.Expectancy') ]
# tmp
(summary(Life.expectancy))
summary(BMI)
```
$SD_x =$ `r round(sd(BMI), 2)`, and 
$SD_y =$ `r round(sd(Life.expectancy), 2)`

### 3) Our Linear Regression Model
```{r}
reg.bmi.le <- lm(Life.expectancy ~ BMI)
plot(BMI,Life.expectancy, main = "Average BMI VS Average Life Expectency")
abline(reg.bmi.le,lwd=2,col="green")
```

##### Linear Regression Coefficients

```{r, echo=FALSE}
# summary(df2)
a_hat <- cov(Life.expectancy,BMI,  use = "pairwise.complete.obs")/var(BMI,na.rm = "True")
# a_hat
b_hat <- mean(Life.expectancy, na.rm = TRUE)-a_hat*mean(BMI, na.rm = TRUE)
# b_hat
y_hat <- a_hat*BMI + b_hat

###optinal editing - The process of verifying the model, we compared the results of the prediction calculated manually to internal function of R.

#plot(BMI,Life.expectancy, main = "y_hat and the oprator abline are the same line")
#abline(reg.bmi.le,lwd=2,col="green")
#lines(BMI, y_hat)
#plot(lm(Life.expectancy ~ BMI))
```

\br
 $\hat{a} =$ \  `r round(a_hat,3)` is the slope of the regression line. In how much will a person's life expectancy change if their BMI increases by one BMI score.  
\n
 $\hat{b} =$ \ `r round(b_hat,2)` is the intercept of the regression line. It is the size of the life expectency that does not depend on BMI. 
Simply put: a person lives 17.19 years first of all, and then depending on his BMI he will live another $\hat{a}$BMI years.

Pearson Correlation Coefficient: $r(X,Y):= \frac{cov(X,Y)}{SD_X SD_Y} =$ `r round(cor(BMI, Life.expectancy, use = "complete.obs"),3)`
\n The correlation between the variables is positive and high.It suggests that the higher the BMI, the higher the life expectancy.

\n R Square Indicator: $R^2:= r^2(X,Y)=$ `r round(cor(BMI, Life.expectancy, use = "complete.obs")^2,3)`

Our average BMI based model explains `r round(cor(BMI, Life.expectancy, use = "complete.obs")^2,3)` of the variance of averagel life expectancy in a country.


**Note**: The connection is not necessarily linear. Becuase in a linear model, as BMI increases, so does life expectancy. In practice, a BMI over 25 describes a situation of "dangerous obesity" and therefore shortens the lifestyle. A parabolic model may be appropriate.

#### Evaluation - Root Mean Square Error (RMSE)
We showed that  $R^2$  is lower than 0.5, a value that is considered low. However, it can be explained by countless other reasons throughout a person's life expectancy, so relatively it is a variable that explains a considerable percentage of variation

```{r, echo=FALSE}
## the first will be the correlation check
# cor(df2$BMI, df2$Life.Expectancy.Years, use = "complete.obs")

## now we will check what about R_squred
#reg.bmi.le <- lm(Life.expectancy ~ BMI)
#summary(reg.bmi.le)


e<-reg.bmi.le$residuals
RMSE<-round(sqrt(sum(e^2)/(length(BMI)-2)),2)
#RMSE

```

```{r , echo=FALSE}
##R_squred is 0.36, great! lets check it with pearson's number
# security_check <- cor(df2$BMI, df2$Life.Expectancy.Years, use = "complete.obs")**2
# security_check
#The R squared score is 0.52, we will confirm the result by increasing the squared Pearson coefficient and making sure they are equal <insert formula here>. Both indices show a positive linear relationship


res <- residuals(reg.bmi.le)
# res
Sorted_Res <- sort(res)
# head(Sorted_Res) # 92   ,    86     ,   142      ,  28     ,    4   ,     33 ,
# tail(Sorted_Res) #      174  ,     145    ,  47   ,   129     , 143  ,     82

# head(Sorted_Res), tail(Sorted_Res)))
# which(x= c(head(Sorted_Res), tail(Sorted_Res)),arr.ind =TRUE)
df2 <- df[-c(92,86,142,28,4,33,174,145,47,129,143,82),]
# df3 <- df2[,]

reg.bmi.le2 <- lm(df2$Life.expectancy ~ df2$BMI)
#plot(BMI,Life.expectancy)
# summary(reg.bmi.le2)
e_new<-reg.bmi.le2$residuals
RMSE_new<-round(sqrt(sum(e_new^2)/(length(df2$BMI)-2)),2)
#RMSE_new

security_check2 <- cor(df2$BMI, df2$Life.expectancy, use = "complete.obs")**2
# security_check2

##The values of the variables were re-examined after omitting the abnormal observations. We believe that standard deviations will decrease


# summary(df)
# summary(df2)
#summary(lm(df2$Life.expectancy ~ df2$BMI))
#summary(lm(Life.expectancy ~ BMI))

# reg.bmi.le <- lm(Life.expectancy ~ BMI)
# plot(BMI,Life.expectancy, main = "Average BMI VS Average Life Expectency")
# abline(reg.bmi.le,lwd=1.5,col="green")
# abline(reg.bmi.le2,lwd=1,col="blue")
```

$RMSE=$ \ `r RMSE` Years of life expectancy, a small variance of the residuals, suggesting a good accuracy of our model.  \br

### Outlires Handeling
Out of the `r length(df$Country)` observations, we removed the 12 values with most extreme residuals.
We omitted 12 extreme residuals.,and the value of RMSE decreased from `r RMSE` to
`r RMSE_new`.

\n $R^2_{new}:= r_{new}^2(X,Y)=$ `r round(security_check2,2)`

Our average BMI based model explains almost `r ceiling (cor(df2$BMI, df2$Life.expectancy)^2)/2` of the variance of average life expectancy in a country.

#### Demonstrating our model with 3 random samples:

```{r , echo=FALSE}
## the first observation in the df 
# "observation number one - afghanistan,2015,65,19.1"
country<-sample(df2$Country, size = 3)
#country

a_hat <- cov(df2$Life.expectancy, df2$BMI)/var(df2$BMI)
df3 <- cbind(df2[df2$Country %in% country,],a_hat)

#a_hat
b_hat <- mean(df3$Life.expectancy, na.rm = TRUE)-a_hat*mean(df3$BMI, na.rm = TRUE)
#b_hat
df3 <- cbind(df3, b_hat)

Y_hat_LEYears <- round(a_hat*df3$BMI + b_hat,2)
Residual <-round(abs(df3$Life.expectancy-Y_hat_LEYears),2)
df4 <- cbind(df3[df3$Country %in% country,], Y_hat_LEYears)
df4 <- cbind(df4, Residual)

head(df4)
```

```{r eval=FALSE, include=FALSE}

# summary(df2)
a_hat <- cov(df2$Life.expectancy, df2$BMI,  use = "pairwise.complete.obs")/var(df2$BMI,na.rm = "True")
a_hat
b_hat <- mean(df2$Life.expectancy, na.rm = TRUE)-a_hat*mean(df2$BMI, na.rm = TRUE)
b_hat
y_hat <- a_hat*df2$BMI + b_hat

example1 <-y_hat[19.1]
residual1 <- example1 - df2$Life.Expectancy.Years[1]
# residual1


##for the next example we will take ob number 103
df2[103,]
example2 <- y_hat[103]
# example2
residual2 <- example2 - df2$Life.expectancy[103]
# residual2
```
