---
title: "Inferential Statistics Project"
author: "Aviv Gelfand and Maor Moshe  -  03/12/2022"
output:
  html_document: 
    fig_height: 4
    df_print: kable
    fig_width: 6
  word_document: default
editor_options:
  chunk_output_type: console
---


#### Recall of our previous project
Recall that in the previews project we used WHO table of Average BMI and life expectancy for 177 countries in the world.
[Source](https://www.kaggle.com/datasets/kumarajarshi/life-expectancy-who?resource=download).

**Predicted / dependent variable Y**: Life expectancy - refers to average number of years a person can expect to live in a country.

**Undependable variable X**: BMI - Body Mass Index units. A calculation using a person's height and weight. 

**We hypothesized** that BMI is an explanatory variable for a person's life expectancy. 


```{r echo=FALSE, message=FALSE, warning=FALSE }
# getwd()
setwd("C:/Users/avivg/OneDrive/Documents/R/Statistical Analysis Course/Lab2") #Aviv's setwd
# setwd("C:/") #Maor's setwd, *aviv* commit it when you working
df <- read.csv("WHO_BMI_2015.csv")
```

### Question 1) Point Estimators
```{r  echo=FALSE, ,message=FALSE, warning=FALSE}
#Q1 #Assume Y distributs norm and asses E(y) and var(y)
# by MLE method:
Y_mu_hat <- mean(Life.expectancy)
# Y_mu_hat

Y_sig_square_hat <- var(Life.expectancy)
# Y_sig_square_hat
# sqrt(Y_sig_square_hat)


#Hanfanim - moment method :
Y_mu_hat_mom <- mean(Life.expectancy)
# Y_mu_hat_mom

Ey_squre <- Y_mu_hat_mom^2
y_squre <- mean(Life.expectancy^2)
Y_sig_squre_hat_mom_MLE <- y_squre - Ey_squre
# Y_sig_squre_hat_mom

```
##### **1. a)** Estimating Expectency and Variance Y
Let's assume that the predicted value, the average life expectancy in the country, is normally distributed. 

We could define an estimator for the expectancy according to the method of MLE or the  moment's method (the same, a result in a normal distribution).

$\hat{\mu_y} := \overline{Y} = \frac{1}{n} \sum_{i = 1}^{n}y_i$ = `r round(Y_mu_hat,2)`


$\hat{\sigma^2_y} := \overline{Y^2} - \overline{Y}^2= \frac{1}{n} \sum_{i = 1}^{n}(y_i-\overline{Y})^2  =$ `r round(Y_sig_squre_hat_mom_MLE,2)`

In addition, we learned in class about a conventional unbiased estimator for the variance of a normal distribution:

$\hat{\sigma^2_y} := S^2_n :=\frac{1}{n-1}\sum_{i = 1}^{n}(y_i-\overline{Y})^2  =$ `r round(var(Life.expectancy),2)`

We will proceed with the conventional estimator $S^2_n$ in the following questions.


```{r echo=FALSE, message=FALSE, warning=FALSE }
m <- min(BMI)
W<- BMI-m
W <-sort(W)
# W
# Moment method 
a_hat_moment <- round((mean(W)^2)/((mean(W^2))-(mean(W)^2)),2)
# a_hat_moment
# mean(BMI)^2
lambda_hat_moment <- round((mean(W))/((mean(W^2))-(mean(W)^2)),2)
# lambda_hat_moment
```

##### **1. b)** Estimating Expectency and Variance of X using Gamma
We denote the independent variable, BMI as $X$, then: 

Let $m=min(X)$ and Let $W:=X-m$ and assume $W \sim Gamma(\alpha, \lambda)$

Then, by the formulas we prooved in class, the point estimator of $\alpha$ would be:
$\hat{\alpha}=\frac{\overline{W}^2}{\overline{W}^2-\overline{W^2}}$ = `r a_hat_moment`

And the point estimator of $\lambda$ would be:
$\hat{\lambda}=\frac{\overline{W}}{\overline{W}^2-\overline{W^2}}$ = `r lambda_hat_moment`

Thus, $W:= X-m \sim Gamma(\alpha =$ `r a_hat_moment` $, \lambda =$ `r lambda_hat_moment` $)$



##### **1. c-d)** Theoretical and Empirical Quantiles [0.1,0.5,0.75,0.9] of X,Y and W: 
```{r echo=FALSE, message=FALSE, warning=FALSE }
#1.c
quantiles <- c(0.1,0.5,0.75,0.9)
quants <- c(0.1,0.5,0.75,0.9)
Y_Theoretical <- qnorm(quantiles, mean = Y_mu_hat_mom, sd = sd(Life.expectancy) )

X_Theoretical <-  qgamma(quantiles, shape = a_hat_moment, scale = lambda_hat_moment) + m

W_Theoretical <- qgamma(quantiles, shape = a_hat_moment, scale = lambda_hat_moment)
# quants_X
#1.d
X_Empirical= quantile(BMI, probs = quantiles) #for x
Y_Empirical = quantile(Life.expectancy, probs=quantiles) # for y
W_Empirical = quantile(W, probs=quants)
library(dplyr)
df1 = data.frame( rbind(Y_Theoretical,Y_Empirical,X_Theoretical,X_Empirical,W_Theoretical,W_Empirical))
colnames(df1) <- c(0.1,0.5,0.75,0.9)
df1
```

We can see a similarity between the theoretical and the empirical parameters.

##### **1. e) Bonus:** Plotting the Theoretical VS Empirical precentiles

```{r echo=FALSE, message=FALSE, warning=FALSE }
library(tidyverse)
library(dplyr)
#1.e hanfanim 2
percentiles <- seq(0.01,0.99,0.01)
# percentiles

quantiles <- seq(0.01,0.99,0.01)
quants <- seq(0.01,0.99,0.01)

Y_Theoretical <- qnorm(quantiles, mean = Y_mu_hat_mom, sd = sd(Life.expectancy) )
# row.names(Y_Theoretical) <- percentiles
X_Theoretical <-  qgamma(quantiles, shape = a_hat_moment, scale = lambda_hat_moment) + m

W_Theoretical <- qgamma(quantiles, shape = a_hat_moment, scale = lambda_hat_moment)
# quants_X
#1.d
X_Empirical= quantile(BMI, probs = quantiles) #for x
Y_Empirical = quantile(Life.expectancy, probs=quantiles) # for y
W_Empirical = quantile(W, probs=quants)


# Y
plot(ecdf(Y_Theoretical),main="Y Values: Theoretical VS Empirical", xlab="Value of Y",	ylab="Precentile", col='lightblue')
lines(ecdf(Y_Empirical), col='darkgrey')
legend("bottomright",pch="-", col=c("lightblue", 'darkgrey'), c("Theoretical", "Empirical"), cex=1,	)

#X
plot(ecdf(X_Theoretical),main="X Values: Theoretical VS Empirical", xlab="Value of X",	ylab="Precentile", col='lightblue')
lines(ecdf(X_Empirical), col='darkgrey')
legend("bottomright",pch="-", col=c("lightblue", 'darkgrey'), c("Theoretical", "Empirical"), cex=1,	)

# W
plot(ecdf(W_Theoretical),main="W Values: Theoretical VS Empirical", xlab="Value of W",	ylab="Precentile", col='lightblue')
lines(ecdf(W_Empirical), col='darkgrey')
legend("bottomright",pch="-", col=c("lightblue", 'darkgrey'), c("Theoretical", "Empirical"), cex=1,	)
```

## Question 2 - Confidence Intervals

#### **CI of E[Y]** where confidence level is 97%

We don't know What $\sigma_Y$ is. Hence, we calculate the confidence interval of an expected life expectancy in a country (our Y) as follows:

$CI = [\overline{Y_n} \pm t_{n-1,1-\frac{a}{2}} \ \frac{S_n}{\sqrt[]{n}}] = [71.82\pm t_{n=177, \ df=0.985} \  \frac{8.07}{ \sqrt[]{177} }] = [70.49, \ 73.15]$


```{r echo=FALSE, message=FALSE, warning=FALSE }
# Question 2 - CI 

## for the expectancy
# mean(Life.expectancy)
# qnorm(0.985)
qt_ = qt(0.985,df=length(Life.expectancy)-1)
Upper_CI_miu_y <- mean(Life.expectancy) + qt_*sd(Life.expectancy)/(sqrt(length(Life.expectancy)))
# Upper_CI_miu_y

Lower_CI_miu_y <-mean(Life.expectancy) - qt_*sd(Life.expectancy)/(sqrt(length(Life.expectancy))) 
# Lower_CI_miu_y
Miu_CI <- c(Lower_CI_miu_y,Upper_CI_miu_y) 
# Miu_CI #70.50 #73.14
# we can see that the value of x_hat is in the range 
# mean(Life.expectancy)
 
## for the varience of life expectancy 
```


#### **CI of var(Y)** where confidence level is 97%

We calculate the confidence interval of the variance of life expectancy in a country (our Y) as follows:

$CI = [\frac{(n-1)S_n^2}{\chi^2_{(n-1),1-\frac{a}{2}}},\frac{(n-1)S_n^2}{\chi^2_{(n-1),\frac{a}{2}}}]=[\frac{(176)*65.25}{\chi^2_{(176), \ 0.96}},\frac{(176)*65.25}{\chi^2_{(176), \ 0.04}}] = [54.95, 78.94]$


```{r echo=FALSE, message=FALSE, warning=FALSE }
n<-length(Life.expectancy)
lower_ci_sigma_y<-(n-1)*var(Life.expectancy)/qchisq(0.955,n-1)
# lower_ci_sigma_y

upper_ci_sigma_y <- 
(n-1)*var(Life.expectancy)/qchisq(0.045,n-1)
# upper_ci_sigma_y
Sigma_CI <- c(lower_ci_sigma_y,upper_ci_sigma_y)
# Sigma_CI #54.94 #78.94
 # we can see that the value of Sy is in the range -->
# var(Life.expectancy)
```

## Question 3 - Hypotheses Testing

We want to examine if there is a correlation between the Average BMI ($X$) and Life Expectancy ($Y$) in a country. 

We divide all of our (X,Y) to: $U = \{(X_i,Y_i)| X_i \ge med(X_i)\}$ and: $L = \{(X_i,Y_i)| X_i < med(X_i)  \}$.

$U$ denotes the **higher BMI world** $:=$ countries where the average person has a higher BMI than the median BMI in the world.

$L$ denotes the **lower BMI world** $:=$ countries where the average person has a lower BMI than the median BMI in the world.

```{r echo=FALSE, message=FALSE, warning=FALSE }
df_low <- df[df$BMI<median(df$BMI),]
df_high <- df[df$BMI>=median(df$BMI),]

# mean(df_low$Life.expectancy)
# mean(df_high$Life.expectancy)

# mean(BMI)
# median(BMI)

y_lower <- df_low$Life.expectancy
y_high <- df_high$Life.expectancy
y_high <- y_high[0:88]
```

#### **3. a)** Verbal formulation of the research hypotheses

Our null hypothesis ($H_0$) assumes no correlation between Life expectancy ($Y$) and Average BMI ($X$). In other words, $H_0$ assumes both of our examined groups are identical, such that the difference of their means is equal to zero.


**Our alternative hypothesis** ($H_1$) is that the **higher BMI world** would have a higher Life Expectancy then the *lower BMI world* (higher BMI where countries are richer).

\br
\br \br
#### **3. b)** Statistical formulation of our hypotheses

Let us denote $D= Y_U-Y_L$ \br

Then: 

  $H_0: \mu_{_D} = 0$    (two-tailed test) 


  $H_1: \mu_{_D} > 0$\br

 \br


```{r echo=FALSE, message=FALSE, warning=FALSE}
d = y_high - y_lower
p_val <- 1-pt(mean(d)/sqrt(var(d)/length(d)), length(d)-1)

#since we assume H1>H0:
# H1 > H0
# p.value <- 1 - pt(q ,df = 87) 
# p.value

# כיוון שערך פי מאוד נמוך. נדחה את הייצ' אפס עבור כל רמת מובהקות גדולה מאפסילון. כלומר, העולם השמן אינו בעל אורך חיים זהה לעולם הרזה בכל רמת מובהקות.
# q_e <- 1 - pt(mean(e)/(sqrt(var(e)/length(e))),87)
# q_e
```
#### **3. c)** Building a statistical test.

Our test-statistic is $\overline D$. We don't know it's real variance then $D \sim t_{88}$ under $H_0$.

Since we don't know the real variance of D, we estimate it as $\sigma_D^2= S_{D}^2=$ `r round(var(d),2)` with the above mentioned formula.

Then: $P_{val}:= p_{_{H_0}}(Reject \ H_0) \underset{\overline D > 0}= 1-pt_{88}(\cfrac{\overline{D}\ -\ 0}{\sqrt{S_n^2/n}})=$ `r p_val`

Hence, let our test be: Reject $H_0$ if $P_{val} > \alpha$ 

 This is a very small p-value, which suggests that the difference in means between the two groups is statistically significant for every $\alpha$. In other words, the data provides strong evidence against $H_0$.

#### **3. d)** Applying our statistical test 

We set our significance level as $\alpha=0.03$ and we achieve:
$P_{val} \approx 0 > \alpha =3 \ \  \rightarrow$ **We Reject** $H_0$ with the significance level of $0.03$.

Based on the p-value, we can conclude that the mean life expectancy of the High BMI" group group is significantly greater then the mean life expectancy of the "Low BMI" group.

**The End**

```{r echo=FALSE, message=FALSE, warning=FALSE }
# **`r round(qt(0.975,87),2)`**
#3.d
alpah = 0.03
p.value = 1.849632e-13
```
<!-- p-value that is less than our significance level, then we can conclude that the null hypothesis is rejected. -->
